COATi: statistical pairwise alignment of protein coding sequences

### 2024-01-16 alignpair_letter.tex
Update supplementary materials.

\section*{Materials and Methods}

@@

@@

\begin{figure}[h!]
    \centering%
    \includegraphics{figures/fig-k2p-empirical.pdf}
    \par
    \caption{COATi.}
    \label{fig:k2p-empirical}
\end{figure}

>>>>

\section*{Methods}

@@

% 8263 gapless genes.
% 5864 gapped genes.

@@

\begin{figure}[h!]
    \centering%
    \includegraphics{figures/fig-k2p-empirical.pdf}
    \par
    \caption{COATi's alignments produce biologically reasonable evolutionary distances. (a) The distribution of K2P distances inferred from alignments generated by each method. The averages of each distribution are indicated by vertical lines. The averages are COATi = 0.0084, ClustalΩ = 0.0178, MACSE = 0.0149, MAFFT = 0.0154, PRANK = 0.0092, and COATi-rev = 0.0084. (b) The distribution of the differences between distances inferred by COATi and other methods. The X-axes of both plots have been pseudo-log transformed using the inverse hyperbolic sine.}
    \label{fig:k2p-empirical}
\end{figure}

% Wilcoxon signed rank test
% coati vs rev-coati: p-value: 0.4212 (two.sided)
% coati vs prank: p-value: < 2.2e-16 (less)
% coati vs macse: p-value: < 2.2e-16 (less)
% coati vs mafft: p-value: < 2.2e-16 (less)
% coati vs clustalo: p-value: < 2.2e-16 (less)

### 2024-01-16 supplementary_materials/supplementary_materials.Rmd
Update supplementary materials.

knitr::opts_chunk$set(echo = FALSE)
```

@@

# Gap Patterns

## Lengths

@@

```{r gap-table-1, results = 'asis', message = FALSE}
gap_data <- read_csv(here("supplementary_materials/data/gap-lengths.csv"))

gap_data <- gap_data |> rename("Gap Type" = "values", "Method" = "method",
  "7+" = "7", "8+" = "8", "9+" = "9")

gap_data1 <- gap_data |> filter(Method %in% c("clustalo", "coati-tri-mg",
    "macse", "mafft", "prank", "rev-coati-tri-mg")) |>
    mutate(Method = fct_relevel(Method, c("coati-tri-mg", "rev-coati-tri-mg"))) |>
    mutate(Method = case_match(Method,
        "clustalo" ~ "ClustalΩ",
        "coati-tri-mg" ~ "COATi",
        "macse" ~ "MACSE",
        "mafft" ~ "MAFFT",
        "prank" ~ "PRANK",
        "rev-coati-tri-mg" ~ "COATi-rev"
        )) |>
    arrange(Method, "Gap Type")

k <- kable(gap_data1, align = "lcrrrrrrrrr", caption="Number of gaps introduced by each method separated by length and type.", booktabs = TRUE, linesep = c("", "\\addlinespace"))

@@

```{r gap-table-2, results = 'asis', message = FALSE}
gap_data2 <- gap_data |> filter(Method %in% c("coati-tri-mg",
    "coati-tri-ecm", "coati-mar-mg", "coati-mar-ecm", "coati-dna")) |>
    mutate(Method = fct_relevel(Method, c("coati-tri-mg",
        "coati-tri-ecm", "coati-mar-mg", "coati-mar-ecm", "coati-dna"))) |>
    arrange(Method, "Gap Type") |>
    rename(Model = Method) |>
    mutate(Model = fct_recode(Model,
        "TRI-MG" = "coati-tri-mg",
        "TRI-ECM" = "coati-tri-ecm",
        "MAR-MG" = "coati-mar-mg",
        "MAR-ECM" = "coati-mar-ecm",
        "DNA" = "coati-dna"
    ))

k <- kable(gap_data2, align = "lcrrrrrrrrr", caption="Number of gaps introduced by different COATi models separated by length and type.", booktabs = TRUE, linesep = c("", "\\addlinespace"))

@@

## Phases

@@

```{r gap-table-3, results = 'asis', message = FALSE}
gap_data <- read_csv(here("supplementary_materials/data/gap-phases.csv"))

gap_data <- gap_data |> rename("Method" = "method")

gap_data1 <- gap_data |> filter(Method %in% c("clustalo", "coati-tri-mg",
    "macse", "mafft", "prank", "rev-coati-tri-mg")) |>
    mutate(Method = fct_relevel(Method, c("coati-tri-mg", "rev-coati-tri-mg"))) |>
    mutate(Method = case_match(Method,
        "clustalo" ~ "ClustalΩ",
        "coati-tri-mg" ~ "COATi",
        "macse" ~ "MACSE",
        "mafft" ~ "MAFFT",
        "prank" ~ "PRANK",
        "rev-coati-tri-mg" ~ "COATi-rev"
        )) |>
    arrange(Method)

k <- kable(gap_data1, align = "lrrr", caption="Number of gaps introduced by each alignment method separated by phase.", booktabs = TRUE, linesep = c("\\addlinespace"))

@@

```{r gap-table-4, results = 'asis', message = FALSE}
gap_data2 <- gap_data |> filter(Method %in% c("coati-tri-mg",
    "coati-tri-ecm", "coati-mar-mg", "coati-mar-ecm", "coati-dna")) |>
    mutate(Method = fct_relevel(Method, c("coati-tri-mg",
        "coati-tri-ecm", "coati-mar-mg", "coati-mar-ecm", "coati-dna"))) |>
    arrange(Method) |>
    rename(Model = Method) |>
    mutate(Model = fct_recode(Model,
        "TRI-MG" = "coati-tri-mg",
        "TRI-ECM" = "coati-tri-ecm",
        "MAR-MG" = "coati-mar-mg",
        "MAR-ECM" = "coati-mar-ecm",
        "DNA" = "coati-dna"
    ))

k <- kable(gap_data2, align = "lrrr", caption="Number of gaps introduced by different COATi models separated by phase.", booktabs = TRUE, linesep = c("\\addlinespace"))

@@

## Gapiness

We quantified the gapiness of different aligners and COATi models (Tab. \@ref(tab:gap-table-5) and \@ref(tab:gap-table-6)). Here we define gapiness as the tendency of an aligner to introduce gaps into an alignment, and we measure it by calculating the percentage of nucleotides aligned with a gap. Note that this is different than the percent of columns that contain a gap.

```{r gap-table-5, results = 'asis', message = FALSE}
gap_data <- read_csv(here("supplementary_materials/data/gapiness.csv"))

gap_data <- gap_data |> rename("Method" = "method") |>
    mutate("Gapiness" = sprintf("%0.2f%%",round(p * 100, 2))) |>
    select(-p)

gap_data1 <- gap_data |>
    filter(Method %in% c("clustalo", "coati-tri-mg",
    "macse", "mafft", "prank", "rev-coati-tri-mg")) |>
    mutate(Method = fct_relevel(Method, c("coati-tri-mg", "rev-coati-tri-mg"))) |>
    mutate(Method = case_match(Method,
        "clustalo" ~ "ClustalΩ",
        "coati-tri-mg" ~ "COATi",
        "macse" ~ "MACSE",
        "mafft" ~ "MAFFT",
        "prank" ~ "PRANK",
        "rev-coati-tri-mg" ~ "COATi-rev"
        )) |>
    arrange(Method) |>
    pivot_wider(names_from = "Method", values_from = "Gapiness")

k <- kable(gap_data1, align = "ccccc", caption="Average gapiness of alignments separated by alignment method.", booktabs = TRUE, linesep = c("\\addlinespace"))

@@

```{r gap-table-6, results = 'asis', message = FALSE}
gap_data2 <- gap_data |> filter(Method %in% c("coati-tri-mg",
    "coati-tri-ecm", "coati-mar-mg", "coati-mar-ecm", "coati-dna")) |>
    mutate(Method = fct_relevel(Method, c("coati-tri-mg",
        "coati-tri-ecm", "coati-mar-mg", "coati-mar-ecm", "coati-dna"))) |>
    arrange(Method) |>
    rename(Model = Method) |>
    mutate(Model = fct_recode(Model,
        "TRI-MG" = "coati-tri-mg",
        "TRI-ECM" = "coati-tri-ecm",
        "MAR-MG" = "coati-mar-mg",
        "MAR-ECM" = "coati-mar-ecm",
        "DNA" = "coati-dna"
    ))  |>
    pivot_wider(names_from = "Model", values_from = "Gapiness")

k <- kable(gap_data2, align = "ccccc", caption="Average gapiness of alignments separated by COATi model.", booktabs = TRUE, linesep = c("\\addlinespace"))

>>>>

cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73",
    "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#999999")
set.seed(17761)

knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(dev = "cairo_pdf")

# Alignment stats
ALN_DATA <- read_csv(here("data", "raw_fasta_aligned_stats.csv.gz"))

# Preprocessed cigar strings from alignment stats
GAP_DATA <- read_csv(here("supplementary_materials", "data", "gaps.csv.gz"))
```

@@

# Empirical Results

## Gap Patterns

### Lengths

@@

```{r gap-table-1, results = 'asis', message = FALSE}
gap_len_data <- GAP_DATA |> filter(op %in% c("I", "D")) |>
    mutate(length = if_else(length <= 6, length, 7 + ((length-1) %% 3))) |>
    arrange(length) |>
    count(method, length, op) |>
    complete(method, op, length = full_seq(length, 1), fill = list(n = 0)) |>
    pivot_wider(names_from = "length", values_from = "n")

gap_len_data <- gap_len_data |> rename("Gap Type" = "op", "Method" = "method",
  "7+" = "7", "8+" = "8", "9+" = "9")

tab <- gap_len_data |> filter(Method %in% c("clustalo", "coati-tri-mg",
    "macse", "mafft", "prank", "rev-coati-tri-mg")) |>
    mutate(Method = fct_relevel(Method, c("coati-tri-mg", "rev-coati-tri-mg"))) |>
    mutate(Method = case_match(Method,
        "clustalo" ~ "ClustalΩ",
        "coati-tri-mg" ~ "COATi",
        "macse" ~ "MACSE",
        "mafft" ~ "MAFFT",
        "prank" ~ "PRANK",
        "rev-coati-tri-mg" ~ "COATi-rev"
        )) |>
    arrange(Method, "Gap Type")

k <- kable(tab, align = "lcrrrrrrrrr", caption="Number of gaps introduced by each method separated by length and type.", booktabs = TRUE, linesep = c("", "\\addlinespace"))

@@

```{r gap-table-2, results = 'asis', message = FALSE}
tab <- gap_len_data |> filter(Method %in% c("coati-tri-mg",
    "coati-tri-ecm", "coati-mar-mg", "coati-mar-ecm", "coati-dna")) |>
    mutate(Method = fct_relevel(Method, c("coati-tri-mg",
        "coati-tri-ecm", "coati-mar-mg", "coati-mar-ecm", "coati-dna"))) |>
    arrange(Method, "Gap Type") |>
    rename(Model = Method) |>
    mutate(Model = fct_recode(Model,
        "TRI-MG" = "coati-tri-mg",
        "TRI-ECM" = "coati-tri-ecm",
        "MAR-MG" = "coati-mar-mg",
        "MAR-ECM" = "coati-mar-ecm",
        "DNA" = "coati-dna"
    ))

k <- kable(tab, align = "lcrrrrrrrrr", caption="Number of gaps introduced by different COATi models separated by length and type.", booktabs = TRUE, linesep = c("", "\\addlinespace"))

@@

### Phases

@@

```{r gap-table-3, results = 'asis', message = FALSE}
gap_phase_data <- GAP_DATA |> 
    mutate(mlen = if_else(op == "I", 0L, length)) |>
    mutate(pos = cumsum(mlen) - mlen, .by=c(gene, method)) |>
    mutate(phase = ((pos-1) %% 3) + 1) |>
    filter(op %in% c("I", "D")) |>
    count(method, phase) |>
    drop_na(phase) |>
    complete(method, phase = full_seq(phase, 1), fill = list(n = 0)) |>
    pivot_wider(names_from = "phase", values_from = "n")

gap_phase_data <- gap_phase_data |> rename("Method" = "method")

tab <- gap_phase_data |> filter(Method %in% c("clustalo", "coati-tri-mg",
    "macse", "mafft", "prank", "rev-coati-tri-mg")) |>
    mutate(Method = fct_relevel(Method, c("coati-tri-mg", "rev-coati-tri-mg"))) |>
    mutate(Method = case_match(Method,
        "clustalo" ~ "ClustalΩ",
        "coati-tri-mg" ~ "COATi",
        "macse" ~ "MACSE",
        "mafft" ~ "MAFFT",
        "prank" ~ "PRANK",
        "rev-coati-tri-mg" ~ "COATi-rev"
        )) |>
    arrange(Method)

k <- kable(tab, align = "lrrr", caption="Number of gaps introduced by each alignment method separated by phase.", booktabs = TRUE, linesep = c("\\addlinespace"))

@@

```{r gap-table-4, results = 'asis', message = FALSE}
tab <- gap_phase_data |> filter(Method %in% c("coati-tri-mg",
    "coati-tri-ecm", "coati-mar-mg", "coati-mar-ecm", "coati-dna")) |>
    mutate(Method = fct_relevel(Method, c("coati-tri-mg",
        "coati-tri-ecm", "coati-mar-mg", "coati-mar-ecm", "coati-dna"))) |>
    arrange(Method) |>
    rename(Model = Method) |>
    mutate(Model = fct_recode(Model,
        "TRI-MG" = "coati-tri-mg",
        "TRI-ECM" = "coati-tri-ecm",
        "MAR-MG" = "coati-mar-mg",
        "MAR-ECM" = "coati-mar-ecm",
        "DNA" = "coati-dna"
    ))

k <- kable(tab, align = "lrrr", caption="Number of gaps introduced by different COATi models separated by phase.", booktabs = TRUE, linesep = c("\\addlinespace"))

@@

## Homology Patterns

We quantified the homology patterns of residues for different aligners and COATi models (Tab. \@ref(tab:gap-table-5), Tab. \@ref(tab:gap-table-6), and Fig. \@ref(fig:gap-figure-1)). Here we define the match, mismatch, and gap percentages as the percent of nucleotides aligned against a match, mismatch, and gap respectively. Note that this is different than the percent of columns that contain a match, mismatch, or gap because match and mismatch columns are counted twice.

```{r gap-table-5, results = 'asis', message = FALSE}
gap_count_data <- GAP_DATA |>
    count(gene, method, op, wt = length) |>
    drop_na(op) |>
    pivot_wider(id_cols=c(gene, method),
        values_from=n, names_from=op,
        values_fill = 0) |>
    rename(A = `=`)

gap_count_data <- gap_count_data |> mutate(
    anc_len = A + X + D,
    dec_len = A + X + I,
    Pa = (A + A)/(anc_len + dec_len),
    Px = (X + X)/(anc_len + dec_len),
    Pg = (D + I)/(anc_len + dec_len),
)

gap_frac_data <- gap_count_data |> 
    summarize(Matches = mean(Pa), Mismatches = mean(Px), Gaps = mean(Pg), .by = method) |>
    mutate(across(Matches:Gaps,  \(x) scales::percent(x, accuracy = 0.01))) |>
    rename(Method = method)

tab <- gap_frac_data |>
    filter(Method %in% c("clustalo", "coati-tri-mg",
    "macse", "mafft", "prank", "rev-coati-tri-mg")) |>
    mutate(Method = fct_relevel(Method, c("coati-tri-mg", "rev-coati-tri-mg"))) |>
    mutate(Method = case_match(Method,
        "clustalo" ~ "ClustalΩ",
        "coati-tri-mg" ~ "COATi",
        "macse" ~ "MACSE",
        "mafft" ~ "MAFFT",
        "prank" ~ "PRANK",
        "rev-coati-tri-mg" ~ "COATi-rev"
        )) |>
    arrange(Method)

k <- kable(tab, align = "rccc", caption="Average homology percentages of alignments separated by alignment method.", booktabs = TRUE, linesep = c("\\addlinespace"))

@@

```{r gap-table-6, results = 'asis', message = FALSE}
tab <- gap_frac_data  |> filter(Method %in% c("coati-tri-mg",
    "coati-tri-ecm", "coati-mar-mg", "coati-mar-ecm", "coati-dna")) |>
    mutate(Method = fct_relevel(Method, c("coati-tri-mg",
        "coati-tri-ecm", "coati-mar-mg", "coati-mar-ecm", "coati-dna"))) |>
    arrange(Method) |>
    rename(Model = Method) |>
    mutate(Model = fct_recode(Model,
        "TRI-MG" = "coati-tri-mg",
        "TRI-ECM" = "coati-tri-ecm",
        "MAR-MG" = "coati-mar-mg",
        "MAR-ECM" = "coati-mar-ecm",
        "DNA" = "coati-dna"
    ))

k <- kable(tab, align = "rccc", caption="Average homology percentages of alignments separated by COATi model.", booktabs = TRUE, linesep = c("\\addlinespace"))

```{r gap-figure-1, fig.cap="Gap fractions of COATi vs other aligners. Each panel is a scatter plot where the x coordinate is the gap fraction of a COATi alignment and y coordinate is the gap fraction of the corresponding alignment from another aligner.", message = FALSE, warning = FALSE}

# Pg

tab <- gap_count_data |>
    filter(method %in% c(c("clustalo", "coati-tri-mg",
    "macse", "mafft", "prank", "rev-coati-tri-mg"))) |>
    pivot_wider(id_cols=gene, names_from=method, values_from=Pg) |>
    rename(coati = `coati-tri-mg`, `coati-rev` = `rev-coati-tri-mg`) |>
    relocate(gene, coati)

tab <- tab |> pivot_longer(cols = clustalo:`coati-rev`,
    names_to="method", values_to="other") |>
    mutate(method = fct_relevel(method, "clustalo", "macse", "mafft",
        "prank", "coati-rev")) |>
    mutate(method = fct_recode(method,
        "vs ClustalΩ" = "clustalo",
        "vs MACSE" = "macse",
        "vs MAFFT" = "mafft",
        "vs PRANK" = "prank",
        "vs COATI-rev" = "coati-rev"
    ))

gg <- ggplot(tab, aes(x = coati, y = other, color = method)) + facet_wrap(vars(method))
gg <- gg + geom_point(alpha = 0.2, show.legend = FALSE)
gg <- gg + xlab("COATi") + ylab("Other")
gg <- gg + scale_color_manual(values = cbbPalette[c(2,4,6,7,8)])
gg <- gg + scale_x_continuous(trans = scales::transform_pseudo_log(0.0003, base=10),
    breaks=c(0, 0.001, 0.01, 0.1, 1), limits = c(0, 1.2))
gg <- gg + scale_y_continuous(trans = scales::transform_pseudo_log(0.0003, base=10),
    breaks=c(0, 0.001, 0.01, 0.1, 1))

gg <- gg + theme_grey(9) + theme(
    panel.grid.minor.x = element_blank(),
    panel.grid.minor.y = element_blank()
)

gg

```

## Evolutionary Distances

We quantified the evolutionary distances inferred by different aligners and COATi models (Fig. \@ref(fig:k2p-figure-1), Tab. \@ref(tab:k2p-table-1), and Tab. \@ref(tab:k2p-table-2)).

```{r k2p-figure-1, fig.cap="COATi produced shorter evolutionary distances than other aligners. Each panel is a scatter plot where the x coordinate is the K2P distance from a COATi alignment and y coordinate is the K2P distance from the corresponding alignment from another aligner.", warning = FALSE}
k2p_data <- ALN_DATA |>
    filter(method %in% c(c("clustalo", "coati-tri-mg",
    "macse", "mafft", "prank", "rev-coati-tri-mg"))) |>
    pivot_wider(id_cols=gene, names_from=method, values_from=k2p) |>
    rename(coati = `coati-tri-mg`, `coati-rev` = `rev-coati-tri-mg`) |>
    relocate(gene, coati)

k2p_data1 <- k2p_data |> pivot_longer(cols = clustalo:`coati-rev`,
    names_to="method", values_to="other") |>
    mutate(method = fct_relevel(method, "clustalo", "macse", "mafft",
        "prank", "coati-rev")) |>
    mutate(method = fct_recode(method,
        "vs ClustalΩ" = "clustalo",
        "vs MACSE" = "macse",
        "vs MAFFT" = "mafft",
        "vs PRANK" = "prank",
        "vs COATI-rev" = "coati-rev"
    ))

gg <- ggplot(k2p_data1, aes(x = coati, y = other, color = method)) + facet_wrap(vars(method))
gg <- gg + geom_point(alpha = 0.2, show.legend = FALSE)
gg <- gg + xlab("COATi") + ylab("Other")
gg <- gg + scale_color_manual(values = cbbPalette[c(2,4,6,7,8)])
gg <- gg + scale_x_continuous(trans = scales::transform_pseudo_log(0.0004, base=10),
    breaks=c(0, 0.001, 0.01, 0.1, 1))
gg <- gg + scale_y_continuous(trans = scales::transform_pseudo_log(0.0004, base=10),
    breaks=c(0, 0.001, 0.01, 0.1, 1))

gg <- gg + theme_grey(9) + theme(
    panel.grid.minor.x = element_blank(),
    panel.grid.minor.y = element_blank()
)

gg
```

```{r k2p-table-1, results = 'asis', message = FALSE}
tab <- ALN_DATA |> summarize(m = mean(k2p, na.rm = TRUE), .by = "method") |>
    rename(Method = method)

tab_data1 <- tab |>
    filter(Method %in% c("clustalo", "coati-tri-mg",
    "macse", "mafft", "prank", "rev-coati-tri-mg")) |>
    mutate(Method = fct_relevel(Method, c("coati-tri-mg", "rev-coati-tri-mg"))) |>
    mutate(Method = case_match(Method,
        "clustalo" ~ "ClustalΩ",
        "coati-tri-mg" ~ "COATi",
        "macse" ~ "MACSE",
        "mafft" ~ "MAFFT",
        "prank" ~ "PRANK",
        "rev-coati-tri-mg" ~ "COATi-rev"
        )) |>
    arrange(Method) |>
    pivot_wider(names_from = "Method", values_from = "m")

k <- kable(tab_data1, digits = 4, align = "ccccc", caption="Average K2P distances of alignments for each method.", booktabs = TRUE, linesep = c("\\addlinespace"), format.args = list(nsmall = 4))

k |> kable_styling(latex_options = "HOLD_position")

```

```{r k2p-table-2, results = 'asis', message = FALSE}
tab_data2 <- tab |> filter(Method %in% c("coati-tri-mg",
    "coati-tri-ecm", "coati-mar-mg", "coati-mar-ecm", "coati-dna")) |>
    mutate(Method = fct_relevel(Method, c("coati-tri-mg",
        "coati-tri-ecm", "coati-mar-mg", "coati-mar-ecm", "coati-dna"))) |>
    arrange(Method) |>
    rename(Model = Method) |>
    mutate(Model = fct_recode(Model,
        "TRI-MG" = "coati-tri-mg",
        "TRI-ECM" = "coati-tri-ecm",
        "MAR-MG" = "coati-mar-mg",
        "MAR-ECM" = "coati-mar-ecm",
        "DNA" = "coati-dna"
    ))  |>
    pivot_wider(names_from = "Model", values_from = "m")

k <- kable(tab_data2, digits = 4, align = "ccccc", caption="Average K2P distances of alignments for each COATi model.", booktabs = TRUE, linesep = c("\\addlinespace"), format.args = list(nsmall = 4))
